Stability of variational quantum dynamics with Gaussian approximants

### 2013-08-21 paper.tex
writing

The state $|α〉$ is a coherent state with complex amplitude $α$.  These can be defined in a number of ways, perhaps most simply in second quantisation, as eigenstates of a boson annihilation operator $a$, with $a|α〉=α|α〉$.  The annihilation operator is not hermitian, so the eigenvalues $α$ are complex numbers.  In first quantisation, such an operator is the complex amplitude
$$a=(2\hbar)^{-½}(λx+ip/λ),$$
where $x$ and $p$ are the position and momentum operators for a particle with mass $m$.  The bosons in this case are vibrational quanta in a notional harmonic potential, centred at the origin, with frequency $ω=λ²/m$.

@@

Before we study the singular values and vectors of $\Gx$, consider why small singular values are a problem.  The pseudoinverse of $\Gx$ with least norm is
$$〈Γ⁺_χ|=∑_{i=1}^∞ v_iσ_i^{-1}〈u_i|.$$  Suppose this is applied to a state $|ψ〉$ that has a low expected particle number, but small components with large particle number.  Such a component will overlap some $|u_i〉$ belonging to small $σ_i$, and $〈Γ⁺_χ|$ will multiply it by $σ_i^{-1}$, generating large components of those $v_i$ in the expansion function.  In the continuous expansion, this isn't a problem: the information is still there, and the state can be recovered by applying $\Gx$ to the expansion.  When the expansion is discretised, however, the small, low particle number component of $〈Γ⁺_χ|ψ〉$ that contain the important information about $|ψ〉$ are very likely to be lost in the large components of $〈Γ⁺_χ|ψ〉$ that give minor details of high particle number components of $|ψ〉$.

@@

\topinsert\XeTeXpicfile converge.pdf width \hsize
Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.  The solid line shows sampling on a regular grid, the dashed one at random amplitudes.  Details in text.
\endinsert

@@

>>>>

The state $|α〉$ is a coherent state with complex amplitude $α$.  These can be defined in a number of ways, perhaps most simply in second quantisation, as eigenstates of a boson annihilation operator $a$, with $a|α〉=α|α〉$.  The annihilation operator is not hermitian, so the eigenvalues $α$ are complex numbers.  In first quantisation, such an operator is the complex amplitude
$$a=(2\hbar)^{-½}(λx+ip/λ),$$
where $x$ and $p$ are the position and momentum operators for a particle with mass $m$.  (The dimensions don't work: $λ²$ is an action, but then $a$ has dimensions.)  The bosons in this case are vibrational quanta in a notional harmonic potential, centred at the origin, with frequency $ω=λ²/m$.

@@

Before we study the singular values and vectors of $\Gx$, consider why small singular values are a problem.  The pseudoinverse of $\Gx$ with least norm is
$$〈Γ⁺_χ|=∑_{i=1}^∞ v_iσ_i^{-1}〈u_i|.$$  Suppose this is applied to a state $|ψ〉$ that has a low expected particle number, but small components with large particle number.  Such a component will overlap some $|u_i〉$ belonging to small $σ_i$, and $〈Γ⁺_χ|$ will multiply it by $σ_i^{-1}$, generating large components of those $v_i$ in the expansion function.  In the continuous expansion, this isn't a problem: the information is still there, and the state can be recovered by applying $\Gx$ to the expansion.  When the expansion is discretised, however, the small, low particle number component of $〈Γ⁺_χ|ψ〉$ that contains the important information about $|ψ〉$ is likely to be lost in the large component of $〈Γ⁺_χ|ψ〉$ that gives minor details of high particle number components of $|ψ〉$.

@@

\topinsert\XeTeXpicfile converge.pdf width \hsize
Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.  The solid line shows sampling on a regular grid, the dashed one at random amplitudes.  Details in text.  The dotted line shows the difference between the expanded state and a vacuum state truncated to the extent of the grid; the norm of the truncation error is around $10^{-3}$.
\endinsert

@@

%% Expansion tension

To capture the concept of a smooth expansion over an irregular set of nonorthogonal vectors, we'll define a quantity called the sampling tension.  Suppose that $|u〉$ and $|v〉$ are nearly parallel vectors, and the state $|ψ〉$ is expanded over a set that includes them, as $|ψ〉≈c_u|u〉+c_v|v〉+⋯$.  A relaxed expansion is one where $c_u|u〉$ and $c_v|v〉$ are as similar as possible; this can be ensured, regardless whether the $c$ are complex numbers of whether $|u〉$ and $|v〉$ are parallel or antiparallel, by requiring that the component of $c_u|u〉$ along $|v〉$ is equal to the component of $c_v|v〉$ along $|u〉$.  The difference between these components is
$$T_{uv}=\left| {c_v〈u|v〉\over ‖|u〉‖²} - {c_u〈v|u〉\over ‖|v〉‖²}\right|.$$
These differences form a tension matrix.  In the case that $|u〉$ and $|v〉$ are nearly orthogonal, their components should be independent.  In this case, the $T_{uv}$ will be small, because of the proportionality to $〈u|v〉$.  The tension matrix is a reasonable measure of the smoothness of an expansion: its Frobenius norm, the sum in quadrature of the components, would be a good summary.
